---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```


## Welcome!

Hey math boffins, welcome to the coding with R class! Let me introduce you to our team:

- Hi, I'm Di! A professor at Monash University, and originally from Wauchope, NSW. 
- This is Earo! She's a PhD student in econometrics and business statistics at Monash University, an expert in methods for temporal data. 
- This is Nick! He's a post-doctoral researcher in econometrics and business statistics at Monash University, and has an R package called naniar.

### Exercise 

*Introduce yourself to the other members of your team, and brainstorm a team name. You've got five minutes!* 

*Nominate a spokesperson  to give everyone a 1 minute intro for the team.*

## Let's get started learning R! 

- *Login to your computer.* 
- *Open RStudio, using `Start -> All Programs`*
- *Create a project for your work. On the very right side of the window is a small blue R, and a drop down menu. Select `New project`, then `New directory`, navigate to the desktop, and name the project `ChooseMATH`. This will create a folder with this name on the desktop. This will be our workspace for the day.*

## What is R!

The R project is a huge open source software collaboration. There is a foundation team of 36 people, and contributions from thousands of people globally. You can learn more about the project by visiting this web site [https://www.r-project.org](https://www.r-project.org).

R is the language of data analysis! It is the top program used for data science today, one of the most powerful tools available, and completely free for anyone to use. 

RStudio is commercial product, but it is free to most of us, and it provides an interface to help us use the R language. 

*If R is the airplane, RStudio is the airport terminal that provides the support.*

## Start coding!

The RStudio window has mutiple panes.

- *Console*: you can type R commands and see output
- *Help*: find out information about functions and packages. This same panel will have tabs for *showing plots* that you make, *view apps* and documents, show *files* in the folder, and *packages* used. 
- *Environment*: lists the objects that you create, such as data sets.
- *Editor*: For creating scripts to run in pieces or as a whole (like this document!)

R is a scripting language, which means that it is just like writing an essay, or a math proof. We write a script to do specific tasks, that we can run again and again, or give to someone else to run. 

*Write the R code required to add two plus two (yes, it is as simple as typing `2+2` below:*

```{r two-plus-two, exercise=TRUE}

```

*R is like a big calculator! Try doing other math operations, like subtraction, multiplication, division, or square root and power operations.*

## Simulation

Let's pretend that we are math teachers and we've just given the class a test. R can be used to simulate results like this. 

- The code `rnorm(35, 70, 10)` will generate a sample of size 35 from a normal model with mean 70 and standard deviation 10. 
- But when you do this you will likely want to save the result into a data object. Lets call this `scores`. 
- To save it you need to assign the result to `scores` using the `gets` operator `<-`. This is just like an `=` sign. 

*Try it, `scores <- rnorm(35, 70, 10)`*

```{r rnorm, exercise=TRUE}

```

Oh, nothing happens. Let's check if we have created a dataset. You can check the length of scores using `length(scores)`. It should be 35. You can also compute the mean and standard deviation of the data using `mean(scores)` and `sd(scores)`


